---
title: "Random forest classification"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(fig.retina = 2,
                      fig.align  = "center")
```


## Parameters

The path to the data folder:

```{r}
data_path <- paste0("/Users/MarcChoisy/Library/CloudStorage/",
                    "OneDrive-OxfordUniversityClinicalResearchUnit/",
                    "GitHub/choisy/typhoid/")
```


## Packages

The required packages:

```{r}
required_packages <- c("dplyr", "purrr", "stringr", "rsample", "glmnet", "glmnetUtils")
```

Making sure that the required packages are installed:

```{r}
to_inst <- required_packages[! required_packages %in% installed.packages()[,"Package"]]
if (length(to_inst)) install.packages(to_inst)
rm(required_packages, to_inst)
```

Loading some of these packages:

```{r message = FALSE}
library(dplyr)
library(purrr)
library(rsample)
library(glmnet)
library(glmnetUtils)
```


## Reading the clean data

The path to the data:

```{r}
nepal <- readRDS(paste0(data_path, "clean_data/nepal.rds"))
```


## Feature engineering

Nepal data set:

```{r}
nepal2 <- nepal |> 
  mutate(across(c(Cough, Diarrhoea, vomiting, Abdopain, Constipation, Headache,
                  Anorexia, Nausea, Typhoid_IgM), ~ .x > 0)) |> 
  na.exclude()
```

```{r}
nepal3 <- select(nepal2, BloodCSResult, Cough:Hepatomegaly)
```

```{r}
nepal3 |> 
  select(where(is.logical)) |> 
  map(table)
```


## Spliting the data

Let's create the train and test data sets:

```{r}
data_split <- initial_split(nepal3)
train_data <- training(data_split)
test_data <- testing(data_split)
```


## Some utilitary functions

```{r}
moving_average <- function(x, w = 20) {
  x |>
    seq_along() |> 
    map(~ tail(c(0, x), -.x)) |> 
    head(-w  + 1) |> 
    map(head, w) |> 
    map_dbl(mean)
}
```

```{r}
plot_smooth <- function(x, w, data = nepal3) {
  data[order(data[[x]]), ] |> 
    select({{ x }}, BloodCSResult) |> 
    map_dfc(moving_average, w) |> 
    plot(col = 4)
}
```

```{r}
confusion_matrix <- function(threshold) {
  nepal5 |> 
    mutate(across(points, ~ .x > threshold)) |> 
    table()
}
```

```{r}
specificity <- function(x) {
  x[1, 1] / sum(x[1, ])
}
```

```{r}
sensitivity <- function(x) {
  x[2, 2] / sum(x[2, ])
}
```

```{r}
accuracy <- function(x) {
  diag(x) / sum(x)
}
```

```{r}
j_index <- function(x) {
  specificity(x) + sensitivity(x) - 1
}
```

A function that converts a named vector of coefficient values into a named vector of
points corresponding to a clinical score:

```{r}
coef2scores <- function(x, n) {
  xmin <- min(x)
  xmax <- max(x)
  stepv <- min(diff(seq(xmin, xmax, le = n)))
  ubrks <- seq(xmin - stepv, xmax + stepv, le = n)
  cbrks <- ubrks - ubrks[which.min(abs(ubrks))]
  setNames(sort(c(-(1:sum(cbrks < 0)), 1:sum(cbrks > 0)))[as.integer(cut(x, cbrks))],
           stringr::str_remove(names(x), "TRUE"))
}
```

A function that computes the points from a data frame `df` of binary valued variable
and a vector `cscore` of clinical scores. The names of the data frame and the the
vector should be the same, although not necessarily in the same order.

```{r}
make_points <- function(df, cscore) {
  colSums(t(as.matrix(df)) * cscore[names(df)])
}
```


## Exploration of the effect of the continuous variables

```{r}
plot_smooth("OralTemperature", 120, nepal3)
plot_smooth("Pulse", 100, nepal3)
plot_smooth("OralTemperature", 120, train_data)
plot_smooth("Pulse", 100, train_data)
```


## The functions

```{r}
recoding <- function(x, temp, pulse) {
  mutate(x, across(OralTemperature, ~ .x > temp),                    # temp
            across(Pulse, ~ .x > pulse))                             # pulse
}
```

```{r}
make_clinical_scores <- function(x, n, penalty, mixture = 0) {
  model <- glmnet(model.matrix(BloodCSResult ~ ., x)[, -1], as.factor(x$BloodCSResult),
                  binomial, alpha = mixture, lambda = penalty)       # penalty, mixture
  coef2scores(coef(model)[-1, ], n)                                  # n
}
```

```{r}
xvj <- function(train, validate, threshold, temp, pulse, n, penalty, mixture = 0) {
  train_recoded <- recoding(train, temp, pulse)                      # temp, pulse
  validate_recoded <- recoding(validate, temp, pulse)                # temp, pulse
  cscore <- make_clinical_scores(train_recoded, n, penalty, mixture) # n, penal, mixt
  tibble(observations  = validate$BloodCSResult,
         points        = make_points(validate_recoded[, -1], cscore)) |> 
    mutate(predictions = points > threshold) |>                      # threshold
    select(-points) |> 
    table() |> 
    j_index()
}
```

```{r}
cv_wrapper <- function(x, threshold, temp, pulse, n, penalty, mixture = 0, o = -9999) {
  map_dbl(x, ~ possibly(xvj, o)(train_data[.x, ], train_data[-.x, ],
                                    threshold, temp, pulse, n, penalty, mixture))
}
```

```{r}
trainings <- map(vfold_cv(train_data)$splits, ~ .x$in_id)
cv_wrapper(trainings, threshold = 7, temp = 38.5, pulse = 125, n = 10, penalty = 1)
```

About 20'

```{r}
the_grid <- expand.grid(threshold = seq(5, 30, 5), temp = seq(36, 40, .5),
                        pulse = seq(75, 150, 25), n = c(5, 10, 15),
                        penalty = exp(seq(-4, 6)))
                        
system.time(
jvals <- pmap(the_grid, function(threshold, temp, pulse, n, penalty)
                          cv_wrapper(trainings, threshold, temp, pulse, n, penalty))
)
```

About 5:22':

```{r}
the_grid2 <- expand.grid(threshold = seq(5, 30, 1), temp = seq(36, 40, .5),
                         pulse = seq(75, 150, 5), n = seq(5, 15, 5),
                         penalty = exp(seq(-4, 6)))

system.time(
jvals2 <- pmap(the_grid2, function(threshold, temp, pulse, n, penalty)
                            cv_wrapper(trainings, threshold, temp, pulse, n, penalty))
)
```


## Another strategy

```{r}
tune_model <- function(formula, data) {
  model <- cva.glmnet(formula, data, family = binomial)  
  lambdas <- map(model$modlist, ~ as_tibble(.x[1:2]))
  alphas <- rep(model$alpha, map_int(lambdas, nrow))
  the_grid <- bind_rows(lambdas)
  the_grid$alpha <- alphas
  the_grid |>
    filter(cvm == min(cvm)) |> 
    select(lambda, alpha) |> 
    unlist()
}
```

```{r}
tuned_coefficients <- function(x, temp, pulse) {
  x_rec <- recoding(x, temp, pulse)
  frmla <- BloodCSResult ~ .
  hyper <- tune_model(frmla, x_rec)
  model <- glmnet(model.matrix(frmla, x_rec)[, -1], as.factor(x_rec$BloodCSResult),
                  binomial, alpha = hyper["alpha"], lambda = hyper["lambda"])
  out <- coef(model)[-1, ]
  setNames(out, stringr::str_remove(names(out), "TRUE"))
}
```

```{r}
min2 <- function(x) {
  sum(map_int(x, min, 0))
}
```

```{r}
max2 <- function(x) {
  sum(map_int(x, max, 0))
}
```

```{r}
a <- tuned_coefficients(train_data, 38.5, 125)
```

```{r}
coef2scores(a, n)
```

Take 20':

```{r}
grid1 <- expand.grid(temp = seq(35, 42, .5), pulse = seq(60, 130, 10)) |> 
  as_tibble() |> 
  mutate(coeffs = map2(temp, pulse, tuned_coefficients, x = train_data))
```

### lskjh

```{r}
n_vals <- 3:15
```

For the first row:

```{r}
row1 <- grid1[1, ]
```

Computing the scores:

```{r}
the_scores <- map(n_vals, coef2scores, x = unlist(row1[["coeffs"]]))
```

For the first score of the first row:

```{r}
score1 <- the_scores[[1]]
```

Looking for the best threshold values:

```{r}
make_thresholds <- function(x) {
  seq(min2(x) + 1, max2(x) - 1)
}
```

```{r}
the_thresholds <- make_thresholds(score1)
```

```{r}
log2fac <- function(x) {
  factor(x, levels = c("FALSE", "TRUE"))
}
```

```{r}
compute_j <- function(test, temp, pulse, score, threshold) {
  tibble(obsv = log2fac(test$BloodCSResult),
         pred = log2fac(make_points(
           recoding(test, temp, pulse)[, -1], score) > threshold)) |> 
    table() |>
    j_index()
}
```

```{r}
test <- train_data
temp <- row1$temp
pulse <- row1$pulse
score <- score1
threshold <- thresholds[1]
```

```{r}
compute_j(train_data, row1$temp, row1$pulse, score1, thresholds[1])
```

```{r}
a <- bootstraps(train_data)
map(a$splits, ~ .$in_id)
```

```{r}

map_dbl(, compute_j, row1$temp, row1$pulse, score1, thresholds[1])
```



### lsqkdjfh

```{r}
make_thresholds <- function(x, n) {
  tmp <- coef2scores(x, n)
  seq(min2(tmp) + 1, max2(tmp) - 1)
}
```

```{r}
out <- map(n_vals, make_thresholds, x = unlist(grid1[1, "coeffs"]))
```


```{r}
compute_j <- function(test, temp, pulse, score, threshold) {
  tibble(obsv = test$BloodCSResult,
         pred = make_points(recoding(test, temp, pulse)[, -1], score) > threshold) |> 
    table() |> 
    j_index()
}
```

```{r}
compute_j(train_data, grid1$temp[1], grid1$pulse[1], the_scores[[1]], )
```


```{r}
a <- unlist(grid1[[1, "coeffs"]])
the_scores <- map(n_vals, coef2scores, x = a)
the_scores[[1]]
```


```{r}
out <- map(n_vals, make_thresholds, x = unlist(grid1[1, "coeffs"]))
tibble(n = rep(n_vals, map_int(out, length)), threshold = unlist(out))
```









