---
title: "Data cleaning"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(message = FALSE)
```


## Parameters

The path to the data folder:

```{r}
data_path <- paste0("/Users/MarcChoisy/Library/CloudStorage/",
                    "OneDrive-OxfordUniversityClinicalResearchUnit/",
                    "GitHub/choisy/typhoid/")
```


## Packages

required packages:

```{r}
required_packages <- c("readxl", "dplyr")
```

Making sure that the required packages are installed:

```{r}
to_inst <- required_packages[! required_packages %in% installed.packages()[,"Package"]]
if (length(to_inst)) install.packages(to_inst)
```

Loading some of these packages:

```{r}
library(readxl)
library(dplyr)
```


## Utilitary function

Tuning `factor()`:

```{r}
factor2 <- function(...) factor(..., ordered = TRUE)
```


## Reading and cleaning data

The files in the raw data folder:

```{r}
files <- dir(paste0(data_path, "raw_data"), full = TRUE)
```

The names of the variables:

```{r}
var_names <- c("culture", "sex", "age", "fever", "cough", "diarrhea", "vomiting",
               "abdominal_pain", "constipation", "headache", "pulse", "temperature",
               "splenomegaly", "hepatomegaly", "WBC", "platelets", "ALT", "IgM", "CRP")
```

CRP levels:

```{r}
crp_levels <- c("<10", "10-40", "40-80", ">80")
```

Reading and cleaning raw data from Nepal:

```{r}
nepal <- files[grep("nepa", files)] |>
  read_excel() |> 
  rename(platelets = Platelets_gro,
         WBC       = WBC_gro) |> 
  select(- contains("gro"), - contains(">"), - contains("<"), - StudyNo, - Study,
         - AST, - DateStudy, - HospitalNo, - `Vomiting...18`, - culture, - Anorexia,
         - Nausea) |> 
  mutate(across(c(Age, Pulse, platelets, ALT, Fever, Cough, Diarrhoea, `Vomiting...26`,
                  Abdopain, Constipation, Headache, scorev1, scorev2, scorev3, scorev4,
                  Score8), as.integer),
         across(c(Splenomegaly, Hepatomegaly),
                ~ .x == "1" | .x == "TRUE" | .x == "Yes"),
         across(Sex, ~ factor(c("female", "male")[(.x == "1" | .x == "Male") + 1])),
         across(Typhoid_IgM, ~ factor2(sub("N", "0", substring(.x, 1, 1)))),
         across(`CRP_mg/L`, ~ factor2(sub(" *\\(.*\\).*$", "", .x),
                                      levels = crp_levels)),
         across(BloodCSResult, ~ .x == "SPA" | .x == "ST")) |> 
  select(BloodCSResult, Sex, Age,
         Fever, Cough, Diarrhoea, `Vomiting...26`, Abdopain, Constipation, Headache,
         Pulse, OralTemperature, Splenomegaly, Hepatomegaly, WBC, platelets, ALT,
         Typhoid_IgM, `CRP_mg/L`, scorev1, scorev2, scorev3, scorev4, Score8) |> 
  setNames(c(var_names, paste0("score", c(1:4, 8))))
```

Reading and cleaning raw data from Cambodia and Bangladesh:

```{r}
cambodia_bangladesh <- files[grep("camb", files)] |>
  read_excel() |> 
  select(culture, sex, Age, feverdays, cough, diarrhoea, vomiting, abdopain,
         constipation, headache, pulse, `temp A`, spleen, hepat, wbc, plts, alt,
         `IgM life assay d0`, `CRP Group...63`, st) |> 
  mutate(across(c(culture, cough, diarrhoea, vomiting, abdopain, constipation,
                  headache, spleen, hepat, ), ~ .x > 0),
         across(sex, ~ factor(c("female", "male")[.x + 1])),
         across(`CRP Group...63`, ~ factor2(crp_levels[.x + 1], levels = crp_levels)),
         across(`IgM life assay d0`, factor2),
         across(st, ~ factor(c("Cambodia", "Bangladesh")[.x])),
         across(c(feverdays, pulse, alt), as.integer)) |> 
  setNames(c(var_names, "country"))
```

Saving to disk:

```{r}
saveRDS(nepal, paste0(data_path, "clean_data/nepal.rds"))
saveRDS(cambodia_bangladesh, paste0(data_path, "clean_data/cambodia_bangladesh.rds"))
```


